---
title: "Assignment-3"
output: 
  html_notebook: 
    highlight: textmate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```
##Reference: Deep Learning with R Text Book#
***

This notebook contains the code samples found in Chapter 6, Section 3 of [Deep Learning with R](https://www.manning.com/books/deep-learning-with-r).

***
##Question 
Use any or all of the methods to improve weather forecasting for the problem we discussed in class.
Don’t forget to eventually run the best-performing models (in terms of validation MAE) on the test set! 
Best three performers over baseline performance of 0.29 get extra points.

## A temperature forecasting problem

Weather timeseries dataset recorded at the Weather Station at the Max-Planck-Institute for Biogeochemistry in Jena, Germany: http://www.bgc-jena.mpg.de/wetter/.

In this dataset, fourteen different quantities (such air temperature, atmospheric pressure, humidity, wind direction, etc.) are recorded every ten minutes, over several years. The original data goes back to 2003, but we limit ourselves to data from 2009-2016. This dataset is perfect for learning to work with numerical timeseries. We will use it to build a model that takes as input some data from the recent past (a few days worth of data points) and predicts the air temperature 24 hours in the future.

Download and uncompress the data as follows:

```{r, eval=FALSE, echo=TRUE}
dir.create("~/Downloads/jena_climate", recursive = TRUE)
download.file(
  "https://s3.amazonaws.com/keras-datasets/jena_climate_2009_2016.csv.zip",
  "~/Downloads/jena_climate/jena_climate_2009_2016.csv.zip"
)
unzip(
  "~/Downloads/jena_climate/jena_climate_2009_2016.csv.zip",
  exdir = "~/Downloads/jena_climate"
)
```


Let's take a look at the data:

```{r, results='hide'}
library(tibble)
library(readr)

data_dir <- "~/Downloads/jena_climate"
fname <- file.path(data_dir, "jena_climate_2009_2016.csv")
data <- read_csv(fname)
```



```{r}
glimpse(data)
```

Here is the plot of temperature (in degrees Celsius) over time. On this plot, you can clearly see the yearly periodicity of temperature.

```{r}
library(ggplot2)
ggplot(data, aes(x = 1:nrow(data), y = `T (degC)`)) + geom_line()
```

On this plot, you can clearly see the yearly periodicity of temperature.

Here is a more narrow plot of the first ten days of temperature data (since the data is recorded every ten minutes, we get 144 data points per day):

```{r}
ggplot(data[1:1440,], aes(x = 1:1440, y = `T (degC)`)) + geom_line()
```

On this plot, you can see daily periodicity, especially evident for the last 4 days. We can also note that this ten-days period must be coming from a fairly cold winter month.

If we were trying to predict average temperature for the next month given a few month of past data, the problem would be easy, due to the reliable year-scale periodicity of the data. But looking at the data over a scale of days, the temperature looks a lot more chaotic. So is this timeseries predictable at a daily scale? Let's find out.

## Preparing the data


The exact formulation of our problem will be the following: given data going as far back as `lookback` timesteps (a timestep is 10 minutes) and sampled every `steps` timesteps, can we predict the temperature in `delay` timesteps?

We will use the following parameter values:

* `lookback = 1440`, i.e. our observations will go back 10 days.
* `steps = 6`, i.e. our observations will be sampled at one data point per hour.
* `delay = 144`, i.e. our targets will be 24 hours in the future.

To get started, we need to do two things:

* Preprocess the data to a format a neural network can ingest. This is easy: the data is already numerical, so we don't need to do any vectorization. However each timeseries in the data is on a different scale (e.g. temperature is typically between -20 and +30, but pressure, measured in mbar, is around 1000). So we will normalize each timeseries independently so that they all take small values on a similar scale.
* Write a generator function that takes the current array of float data and yields batches of data from the recent past, along with a target temperature in the future. Because the samples in the dataset are highly redundant (sample _N_ and sample _N + 1_ will have most of their timesteps in common), it would be wasteful to explicitly allocate every sample. Instead, you'll generate the samples on the fly using the original data.

First, you'll convert the R data frame which we read earlier into a matrix of floating point values (we'll discard the first column which included a text timestamp):

```{r}
data <- data.matrix(data[,-1])
```

You'll then preprocess the data by subtracting the mean of each timeseries and dividing by the standard deviation. You're going to use the first 200,000 timesteps as training data, so compute the mean and standard deviation for normalization only on this fraction of the data.
#Reduced the sample set to 100000 when compared to original analysis in textbook
```{r}
train_data <- data[1:100000,]
mean <- apply(train_data, 2, mean)
std <- apply(train_data, 2, sd)
data <- scale(data, center = mean, scale = std)
```

Now here is the data generator you'll use. It yields a list `(samples, targets)`, where `samples` is one batch of input data and `targets` is the corresponding array of target temperatures. It takes the following arguments:

* `data` -- The original array of floating-point data, which you normalized in listing 6.32.
* `lookback` -- How many timesteps back the input data should go.
* `delay` -- How many timesteps in the future the target should be.
* `min_index` and `max_index` -- Indices in the `data` array that delimit which timesteps to draw from. This is useful for keeping a segment of the data for validation and another for testing.
* `shuffle` -- Whether to shuffle the samples or draw them in chronological order.
* `batch_size` -- The number of samples per batch.
* `step` -- The period, in timesteps, at which you sample data. You'll set it 6 in order to draw one data point every hour.

```{r}
generator <- function(data, lookback, delay, min_index, max_index,
                      shuffle = FALSE, batch_size = 128, step = 6) {
  if (is.null(max_index))
    max_index <- nrow(data) - delay - 1
  i <- min_index + lookback
  function() {
    if (shuffle) {
      rows <- sample(c((min_index+lookback):max_index), size = batch_size)
    } else {
      if (i + batch_size >= max_index)
        i <<- min_index + lookback
      rows <- c(i:min(i+batch_size-1, max_index))
      i <<- i + length(rows)
    }
    
    samples <- array(0, dim = c(length(rows), 
                                lookback / step,
                                dim(data)[[-1]]))
    targets <- array(0, dim = c(length(rows)))
                     
    for (j in 1:length(rows)) {
      indices <- seq(rows[[j]] - lookback, rows[[j]] - 1, 
                     length.out = dim(samples)[[2]])
      samples[j,,] <- data[indices,]
      targets[[j]] <- data[rows[[j]] + delay,2]
    }            
    
    list(samples, targets)
  }
}
```

The `i` variable contains the state that tracks next window of data to return, so it is updated using
superassignment (e.g. `i <<- i + length(rows)`).

#Now, let's use the abstract `generator` function to instantiate three generators: one for training, one for validation, and one for testing. Each will look at different temporal segments of the original data: the training generator looks at the first 100,000 timesteps, the validation generator looks at the following 50,000, and the test generator looks at the remainder.

# Trying to reduce the sample to see if it works in regular laptop

```{r}
lookback <- 1440
step <- 6
delay <- 144
batch_size <- 128

train_gen <- generator(
  data,
  lookback = lookback,
  delay = delay,
  min_index = 1,
  max_index = 100000,
  shuffle = TRUE,
  step = step, 
  batch_size = batch_size
)

val_gen = generator(
  data,
  lookback = lookback,
  delay = delay,
  min_index = 100001,
  max_index = 150000,
  step = step,
  batch_size = batch_size
)

test_gen <- generator(
  data,
  lookback = lookback,
  delay = delay,
  min_index = 200001,
  max_index = NULL,
  step = step,
  batch_size = batch_size
)

# This is how many steps to draw from `val_gen`
# in order to see the whole validation set:
val_steps <- (200001 - 100001 - lookback) / batch_size

  # This is how many steps to draw from `test_gen`
# in order to see the whole test set:
test_steps <- (nrow(data) - 200001 - lookback) / batch_size
```


## basic deep learning

In the same way that it's useful to establish a common-sense baseline before trying machine-learning approaches, it's useful to try simple, cheap machine-learning models (such as small, densely connected networks) before looking into complicated and computationally expensive models such as RNNs. This is the best way to make sure any further complexity you throw at the problem is legitimate and delivers real benefits.

The following listing shows a fully connected model that starts by flattening the data and then runs it through two dense layers. Note the lack of activation function on the last dense layer, which is typical for a regression problem. You use MAE as the loss. Because you're evaluating on the exact same data and with the exact same metric you did with the common-sense approach, the results will be directly comparable.
#Approach1 : Adjusting the rmsprop learning rate to 0.001
```{r, echo=TRUE, results='hide'}
library(keras)

model <- keras_model_sequential() %>% 
  layer_flatten(input_shape = c(lookback / step, dim(data)[-1])) %>% 
  layer_dense(units = 32, activation = "relu") %>% 
  layer_dense(units = 1)

model %>% compile(
  optimizer = optimizer_rmsprop(lr = 0.001),
  loss = "mae"
)

history <- model %>% fit_generator(
  train_gen,
  steps_per_epoch = 500,
  epochs = 20,
  validation_data = val_gen,
  validation_steps = val_steps
)
```

Results:

```{r}
plot(history)
```


## recurrent neural network

#Approach2 : Instead of the GRU layer introduced in the textbook model,changed to the LSTM layer.

```{r echo=TRUE, results='hide'}
model <- keras_model_sequential() %>% 
  layer_lstm(units = 32, input_shape = list(NULL, dim(data)[[-1]])) %>% 
  layer_dense(units = 1)

model %>% compile(
  optimizer = optimizer_rmsprop(),
  loss = "mae"
)

history <- model %>% fit_generator(
  train_gen,
  steps_per_epoch = 500,
  epochs = 20,
  validation_data = val_gen,
  validation_steps = val_steps
)
```

Results:

```{r}
plot(history)
```

## Stacking recurrent layers


Increasing network capacity is typically done by increasing the number of units in the layers, or adding more layers. Recurrent layer stacking is a classic way to build more powerful recurrent networks.

To stack recurrent layers on top of each other in Keras, all intermediate layers should return their full sequence of outputs (a 3D tensor) rather than their output at the last timestep. This is done by specifying `return_sequences = TRUE`: 

# Approach3 :Here the units in first layer_gru is increased to 64 and second layer_gru is increased to 128.

```{r, echo=TRUE, results='hide'}
model <- keras_model_sequential() %>% 
  layer_gru(units = 64, 
            dropout = 0.1, 
            recurrent_dropout = 0.5,
            return_sequences = TRUE,
            input_shape = list(NULL, dim(data)[[-1]])) %>% 
  layer_gru(units = 128, activation = "relu",
            dropout = 0.1,
            recurrent_dropout = 0.5) %>% 
  layer_dense(units = 1)

model %>% compile(
  optimizer = optimizer_rmsprop(),
  loss = "mae"
)

history <- model %>% fit_generator(
  train_gen,
  steps_per_epoch = 500,
  epochs = 40,
  validation_data = val_gen,
  validation_steps = val_steps
)
```

Results:

```{r}
plot(history)
```
Summary

#Notes

Initially step is to import data and plotted temperature over time.
Prepare the data: Only 100000 as training data Scaling the data using mean and standard deviation.Define data generator function.Apply generator function training samples of 100000 and validation samples of 50000.

#Approaches planning to use to run to improve performamce 
1:  Adjusting the rmsprop learning rate to 0.001
2:  Instead of the GRU layer introduced in the textbook model,changed to the LSTM layer.
3:  The units in first layer_gru is increased to 64 and second layer_gru is increased to 128 in Stacked Recurrent Model.

To find the best performance over the baseline 0.29 (baseline model:training samples of 200000 and validation samples of 100000) plan to run above mentioned 3 approaches code initially by reducing the sample sizes due to some issues with my computational power unable to verify performance (accuarcy,loss).



